---
title: "Final_896_STATs"
output: html_notebook
---

```{r}
# 1. Stack our finals and add a year column
all_years <- bind_rows(
  final2018 %>% mutate(year = 2018),
  final2019 %>% mutate(year = 2019),
  final2020 %>% mutate(year = 2020),
  final2021 %>% mutate(year = 2021),
  final2022 %>% mutate(year = 2022)
)
```

```{r}
glimpse(all_years)
```

```{r}
# Make sure both data frames have the same county spelling
pa_food5_tidy <- pa_food5 %>%
  mutate(county = sub(" County$", "", county))  # strip “ County”

all_years <- all_years %>%
  left_join(
    pa_food5_tidy %>% select(county, pct_li_la),
    by = "county"
  )

```

```{r}
all_years %>%
  distinct(county, year, pct_li_la) %>%
  arrange(county, year) %>%
  print(n = Inf)

```
```{r}
# first clean up pa_food5 to match our `county` values
pa_food5_clean <- pa_food5 %>%
  mutate(
    county = str_remove(county, " County$")   # drop the word “ County”
  ) %>%
  select(county, pct_li_la)

# then join on county
all_years <- all_years %>%
  left_join(pa_food5_clean, by = "county")

```


```{r}
all_years %>% distinct(county, pct_li_la) 

```
```{r}
all_years <- all_years %>%
  mutate(
    pct_li_la = if_else(year == 2019, pct_li_la, NA_real_)
  )

```

```{r}
glimpse(all_years)
```

```{r}
library(dplyr)
library(stringr)

# 1. Clean up food‐access table
pa_food5_clean <- pa_food5 %>%
  mutate(
    county = str_remove(county, " County$")  # drop the word “ County”
  ) %>%
  select(county, pct_li_la)

# 2. Re‑join it once to our panel
all_years <- all_years %>%
  left_join(pa_food5_clean, by = "county")

# 3. Verify
all_years %>% 
  filter(county == "Centre") %>% 
  select(year, pct_li_la.y) %>% 
  print(n = 5)


```

```{r}
library(dplyr)
library(plm)

# 1. Make sure we only got one pct_li_la per county
pa_food5_clean <- pa_food5 %>%
  mutate(county = str_remove(county, " County$")) %>%    # drop “ County”
  distinct(county, pct_li_la)

```

```{r}
# 2. Join it onto our panel
panel <- all_years %>%
  left_join(pa_food5_clean, by = "county")

# 3. Run a simple pooled OLS (time‑invariant will stay)
pooled <- lm(
  mortality_rate ~ hardship_index_5
                 + percent_food_insecure
                 + pct_broadband
```


```{r}
+ pct_li_la,
  data = panel
)
summary(pooled)

```

```{r}

mundlak <- plm(
  mortality_rate ~ hardship_index_5
                 + percent_food_insecure
                 + pct_broadband
                 + pct_li_la                # between effect
                 + I(pct_li_la - ave(pct_li_la, county))  # within deviation (will be zero, so captures only between)
                 ,
  data  = panel,
  index = c("county","year"),
  model = "within"
)
summary(mundlak)
```



```{r}
# 2. Summarize by year AND county
county_year_summary <- all_years %>%
  group_by(year, county) %>%
  summarize(
    mean_hardship = mean(hardship_index_5, na.rm = TRUE),
    mean_broad    = mean(pct_broadband,    na.rm = TRUE),
    mean_mort     = mean(mortality_rate,   na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(county, year)

print(county_year_summary)


```

#2. Descriptive trends

Time‐series plots of key variables by county

```{r}
ggplot(all_years, aes(x=year, y=hardship_index_5, color=county)) + geom_line() + geom_point()
ggplot(all_years, aes(x=year, y=pct_broadband, color=county)) + geom_line() + geom_point()
ggplot(all_years, aes(x=year, y=mortality_rate, color=county)) + geom_line() + geom_point()

```

#→ Spot which counties saw rising hardship or broadband access, and whether mortality moved in step

================================================================================================

# 2.0 Summary tables by year

```{r}
all_years %>%
  group_by(year, county) %>%
  summarize(
    mean_hardship = mean(hardship_index_5, na.rm = TRUE),
    mean_broad    = mean(pct_broadband,    na.rm = TRUE),
    mean_mort     = mean(mortality_rate,   na.rm = TRUE)
  ) %>%
  arrange(year, county)

```

```{r}
county_year_summary %>%
  pivot_wider(
    names_from   = year,
    values_from  = c(mean_hardship, mean_broad, mean_mort),
    names_sep    = "_"
  )

```

```{r}
all_years %>%
  group_by(year, county) %>%
  summarize(
    mean_hardship = mean(hardship_index_5, na.rm = TRUE),
    mean_broad    = mean(pct_broadband,    na.rm = TRUE),
    mean_mort     = mean(mortality_rate,   na.rm = TRUE),
    .groups       = "drop"
  ) %>%
  arrange(county, year) %>%
  pivot_wider(
    names_from  = year,
    values_from = c(mean_hardship, mean_broad, mean_mort),
    names_sep   = "_"
  )

```

```{r}
install.packages("xfun")

```

```{r}
install.packages(c("gt","tidyverse"))

```


```{r}
library(gt)

# 1. Build the summary
county_year_summary <- all_years %>%
  group_by(year, county) %>%
  summarize(
    Hardship  = mean(hardship_index_5, na.rm=TRUE),
    Broadband = mean(pct_broadband,    na.rm=TRUE),
    Mortality = mean(mortality_rate,   na.rm=TRUE),
    .groups    = "drop"
  )

# 2. Render with gt
county_year_summary %>%
  gt(groupname_col = "year") %>%
  fmt_number(
    columns  = c(Hardship, Broadband, Mortality),
    decimals = 2
  ) %>%
  cols_label(
    year      = "Year",
    county    = "County",
    Hardship  = "Mean Hardship Index",
    Broadband = "Mean % Broadband Access",
    Mortality = "Mean Mortality Rate"
  ) %>%
  tab_header(
    title    = md("**County‐Year Summary (2018–2022)**"),
    subtitle = "Five Pennsylvania Counties"
  ) %>%
  tab_source_note(
    source_note = "Data: CHR (social & broadband) + County mortality"
  )

```

```{r}
library(ggplot2)
library(patchwork)

p1 <- ggplot(all_years, aes(x=year, y=hardship_index_5, color=county)) +
  geom_line(size=1) + geom_point() +
  labs(title="Hardship Index", y="Hardship") +
  theme_minimal() +
  theme(legend.position="none")

p2 <- ggplot(all_years, aes(x=year, y=pct_broadband, color=county)) +
  geom_line(size=1) + geom_point() +
  labs(title="% Broadband Access", y="% Access") +
  theme_minimal() +
  theme(legend.position="none")

p3 <- ggplot(all_years, aes(x=year, y=mortality_rate, color=county)) +
  geom_line(size=1) + geom_point() +
  labs(title="Mortality Rate", y="Deaths per 100k") +
  theme_minimal() +
  theme(legend.position="bottom")

(p1 | p2) / p3 +
  plot_annotation(
    title = "Trends in Hardship, Broadband & Mortality (2018–2022)",
    subtitle = "Five Pennsylvania Counties",
    theme = theme(plot.title = element_text(size=14, face="bold"))
  )

```

```{r}
library(reactable)
reactable(
  county_year_summary,
  columns = list(
    Hardship = colDef(format = colFormat(digits = 2)),
    Broadband= colDef(format = colFormat(digits = 2, suffix = "%")),
    Mortality= colDef(format = colFormat(digits = 1))
  ),
  groupBy = "year",
  defaultPageSize = 10
)

```
=========================================================================================================


```{r}

library(dplyr)
library(tidyr)
library(ggplot2)
library(plm)

```


```{r}
all_years <- all_years %>%
  mutate(period = if_else(year < 2020, "pre_covid", "post_covid"))

all_years %>%
  group_by(period) %>%
  summarize(
    avg_hardship = mean(hardship_index_5, na.rm=TRUE),
    avg_foodinsec = mean(percent_food_insecure, na.rm=TRUE),
    avg_broadband = mean(pct_broadband, na.rm=TRUE),
    avg_mortality = mean(mortality_rate, na.rm=TRUE)
  )

```

```{r}
all_years %>%
  group_by(period) %>%
  summarize(
    cor_pov    = cor(hardship_index_5, mortality_rate, use="complete.obs"),
    cor_food   = cor(percent_food_insecure, mortality_rate, use="complete.obs"),
    cor_bb     = cor(pct_broadband, mortality_rate, use="complete.obs")
  )

```
#2. Panel Regression: County Fixed Effects
--within‐county (fixed‐effects) model isolates how changes in poverty, food insecurity, and broadband within --each county predict mortality, controlling for all time‑invariant county traits:
```{r}
library(plm)

# 2a. Base FE model
fe_base <- plm(
  mortality_rate ~ hardship_index_5 + percent_food_insecure + pct_broadband,
  data  = all_years,
  index = c("county","year"),
  model = "within"
)
summary(fe_base)

```

#3. Testing Pre‑ vs. Post‑COVID Shifts

A. Interaction with a COVID dummy
```{r}
all_years <- all_years %>%
  mutate(post = if_else(year >= 2020, 1, 0))

fe_interact <- plm(
  mortality_rate ~ 
    hardship_index_5*post +
    percent_food_insecure*post +
    pct_broadband*post,
  data  = all_years,
  index = c("county","year"),
  model = "within"
)
summary(fe_interact)

```

#B. Separate FE models pre & post
Compare the size and significance of your predictors across the two periods.
```{r}
fe_pre  <- plm(mortality_rate ~ hardship_index_5 + percent_food_insecure + pct_broadband,
               data = filter(all_years, year < 2020),
               index=c("county","year"), model="within")
fe_post <- plm(mortality_rate ~ hardship_index_5 + percent_food_insecure + pct_broadband,
               data = filter(all_years, year >= 2020),
               index=c("county","year"), model="within")
summary(fe_pre); summary(fe_post)

```

```{r}
library(lmtest)
library(sandwich)
coeftest(fe_interact,
         vcov = vcovHC(fe_interact, type="HC1", cluster="group"))

```

#Add time fixed effects to soak up common shocks (e.g. a national pandemic wave):
```{r}
fe_time <- plm(
  mortality_rate ~ hardship_index_5*post + percent_food_insecure*post + pct_broadband*post + factor(year),
  data  = all_years,
  index = c("county","year"),
  model = "within"
)
summary(fe_time)

```

# 
```{r}
re_model <- plm( mortality_rate ~ hardship_index_5 + percent_food_insecure + pct_broadband,
                 data=all_years, model="random", index=c("county","year"))
phtest(fe_base, re_model)

```

```{r}
library(margins)
m <- lm(mortality_rate ~ hardship_index_5*post + percent_food_insecure*post + pct_broadband*post + factor(county),
        data=all_years)
plot(margins(m), pars = c("hardship_index_5","hardship_index_5:post"))

```



```{r}
# Period‐specific correlations
all_years %>%
  group_by(period) %>%
  summarize(
    cor_pov  = cor(hardship_index_5,    mortality_rate, use="complete.obs"),
    cor_food = cor(percent_food_insecure, mortality_rate, use="complete.obs"),
    cor_bb   = cor(pct_broadband,       mortality_rate, use="complete.obs"),
    .groups  = "drop"
  )

# Fixed‐effects panel regression with COVID interaction
all_years <- all_years %>%
  mutate(post = if_else(year >= 2020, 1, 0))

fe_interact <- plm(
  mortality_rate ~ 
    hardship_index_5*post +
    percent_food_insecure*post +
    pct_broadband*post,
  data  = all_years,
  index = c("county","year"),
  model = "within"
)
summary(fe_interact)

```

```{r}
library(dplyr)
library(plm)

# 1a. Create one‑year lags for our key predictors
all_years_lag <- all_years %>%
  arrange(county, year) %>%
  group_by(county) %>%
  mutate(
    lag_food  = lag(percent_food_insecure, 1),
    lag_pov   = lag(hardship_index_5,      1),
    lag_bb    = lag(pct_broadband,         1)
  ) %>%
  ungroup()

# 1b. Run a fixed‑effects model with lagged predictors and COVID interaction
fe_lag <- plm(
  mortality_rate ~ 
    lag_food*post + lag_pov*post + lag_bb*post,
  data  = all_years_lag,
  index = c("county","year"),
  model = "within"
)
summary(fe_lag)

```

```{r}
library(plm)

# 3a. Create year‑centered dummies (excluding 2019 as our base)
all_years <- all_years %>%
  mutate(across(year, as.factor))

# 3b. FE event‑study model
es_model <- plm(
  mortality_rate ~ 
    # interactions of interest
    hardship_index_5:factor(year) +
    percent_food_insecure:factor(year) +
    pct_broadband:factor(year) +
    # main effects of controls or fixed county effects are implicit
    factor(year),
  data  = all_years,
  index = c("county","year"),
  model = "within"
)

# 3c. Extract and plot coefficients
library(broom)
coefs <- tidy(es_model) %>%
  filter(str_detect(term, "hardship_index_5:factor\\(year\\)")) %>%
  mutate(
    year = str_extract(term, "\\d{4}") %>% as.integer(),
    estimate = estimate,
    conf_low  = estimate - 1.96*std.error,
    conf_high = estimate + 1.96*std.error
  )

ggplot(coefs, aes(x=year, y=estimate)) +
  geom_hline(yintercept = 0, linetype="dashed") +
  geom_pointrange(aes(ymin=conf_low, ymax=conf_high)) +
  labs(
    title = "Event‐Study: Hardship → Mortality by Year",
    x     = "Year",
    y     = "Slope on Hardship Index"
  ) +
  theme_minimal()

```

```{r}
library(dplyr)
library(sf)
library(tigris)
library(ggplot2)
options(tigris_use_cache = TRUE)

```


```{r}
# 1. Pull in the county shapes
pa_counties <- counties(state = "PA", cb = TRUE, class = "sf")

# 2. Filter to your five counties and join
map_df <- pa_counties %>%
  filter(NAME %in% c("Centre","Clinton","Huntingdon","Juniata","Mifflin")) %>%
  left_join(all_years, by = c("NAME" = "county"))

# 3. Now we can ggplot safely:
ggplot(data = map_df) +
  geom_sf(aes(fill = hardship_index_5), color = "white") +
  facet_wrap(~ year, ncol = 3) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    fill    = "Hardship Index",
    title   = "County Hardship Index (2018–2022)",
    caption = "Source: CHR + County mortality"
  ) +
  theme_void()

```
```{r}
# 1. Pull in the county shapes
pa_counties <- counties(state = "PA", cb = TRUE, class = "sf")

# 2. Filter to our five counties and join
map_df <- pa_counties %>%
  filter(NAME %in% c("Centre","Clinton","Huntingdon","Juniata","Mifflin")) %>%
  left_join(all_years, by = c("NAME" = "county"))

# 3. Plot hardship with county labels
ggplot(data = map_df) +
  geom_sf(aes(fill = hardship_index_5), color = "white") +
  geom_sf_text(aes(label = NAME), size = 3, color = "black") +
  facet_wrap(~ year, ncol = 3) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    fill    = "Hardship Index",
    title   = "County Hardship Index (2018–2022)",
    caption = "Source: CHR + County mortality"
  ) +
  theme_void()
```
===========================================================================


```{r}
library(dplyr)
library(knitr)
library(tidyr)


demo_bb_2021 %>%
  transmute(
    County             = county,
    `Population (est)` = population,
    `% Rural`          = pct_rural,
    `HS Grad+ (%)`     = hs_grad,
    `Bachelor’s+ (%)`  = bach_plus,
    `Median HH Income` = med_hh_inc,
    `% ≥3 Broadband`   = pct_bb3
  ) %>%
  kable(
    caption = "Table 1. Central PA County Characteristics, 2021",
    digits  = c(0, 1, 1, 1, 1, 0, 1)
  )

```

```{r}
library(dplyr)
library(tidyr)  

df_panel <- df_panel %>%
  left_join(demo_bb_2021 %>% mutate(year = 2021),
            by = c("county","year")) %>%
  arrange(county, year) %>%
  group_by(county) %>%
  fill(population:pct_bb3, .direction = "down") %>%  # carry 2021 values forward
  fill(population:pct_bb3, .direction = "up") %>%    # then backward
  ungroup()

```

```{r}
df_panel %>% filter(year == 2021) %>% select(county, population:pct_bb3)

```
#Table A:
```{r}
library(dplyr)
library(knitr)

# Our 2021 lookup (demo_bb_2021) already has exactly what we need:
demo_bb_2021 %>%
  transmute(
    County             = county,
    `Population (est)` = population,
    `% Rural`          = pct_rural,
    `HS Grad+ (%)`     = hs_grad,
    `Bachelor’s+ (%)`  = bach_plus,
    `Median HH Income` = med_hh_inc,
    `% ≥3 Broadband`   = pct_bb3
  ) %>%
  kable(
    caption = "Table 1. Central PA County Characteristics, 2021",
    digits  = c(0, 1, 1, 1, 1, 0, 1)
  )

```

```{r}
library(ggplot2)
library(viridis)    # for scale_fill_viridis_c()
library(tigris)
library(sf)

options(tigris_use_cache = TRUE)

# 1. Load PA county geometries and filter
pa_cty <- counties(state="PA", year=2021, class="sf") %>%
  rename(county = NAME) %>%
  filter(county %in% c("Centre","Clinton","Huntingdon","Juniata","Mifflin"))

# 2. Join your hardship panel
map_df <- pa_cty %>%
  left_join(df_panel, by = "county")

# 3. Plot
ggplot(map_df) +
  geom_sf(aes(fill = hardship5), color = "white", size = 0.2) +
  geom_sf_text(aes(label = county), size = 3, color = "black") +
  scale_fill_viridis_c(
    option    = "magma",
    direction = -1,
    name      = "Hardship\nIndex (z)",
    na.value  = "grey90"
  ) +
  facet_wrap(~ year, ncol = 3) +
  coord_sf(datum = NA) +                              # drop default graticule labels
  theme_void(base_size = 12) +                        # starts from a blank canvas
  theme(
    strip.background = element_rect(fill = "white", colour = "black"),
    strip.text       = element_text(face = "bold", size = 12),
    legend.position  = "right",
    legend.title     = element_text(size = 10),
    legend.text      = element_text(size = 8),
    plot.title       = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle    = element_text(size = 10, hjust = 0.5),
    plot.caption     = element_text(size = 8, hjust = 1)
  ) +
  labs(
    title    = "Figure 1. Five‑Variable Hardship Index by County (2018–2022)",
    subtitle = "Higher z‑scores indicate greater socioeconomic + food‑access stress",
    caption  = "Source: County Health Rankings · USDA · FCC"
  )


```

```{r}
ggplot(map_df) +
  geom_sf(aes(fill = hardship5), color = "white", size = 0.2) +
  geom_sf_text(aes(label = county), size = 3) +
  scale_fill_viridis_c(
    option    = "viridis",    # default, but a bit lighter than "magma"
    direction = 1,            # low → high maps to light → dark
    name      = "Hardship\nIndex (z)",
    na.value  = "grey90"
  ) +
  facet_wrap(~ year, ncol = 3) +
  coord_sf(datum = NA) +
  theme_void() +
  theme(
    strip.background = element_rect(fill = "white", colour = "black"),
    strip.text       = element_text(face = "bold", size = 12),
    legend.position  = "right",
    plot.title       = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle    = element_text(size = 10, hjust = 0.5),
    plot.caption     = element_text(size = 8, hjust = 1)
  ) +
  labs(
    title    = "Figure 1. Five‑Variable Hardship Index by County (2018–2022)",
    subtitle = "Higher z‑scores indicate greater socioeconomic + food‑access stress",
    caption  = "Source: CHR · USDA · FCC"
  )

```

```{r}
library(RColorBrewer)

ggplot(map_df) +
  geom_sf(aes(fill = hardship5), color = "white", size = 0.2) +
  geom_sf_text(aes(label = county), size = 3) +
  scale_fill_distiller(
    palette   = "YlGnBu",    # yellow–green–blue
    direction = 1,           # low = light yellow, high = dark blue
    name      = "Hardship\nIndex (z)",
    na.value  = "grey90"
  ) +
  facet_wrap(~ year, ncol = 3) +
  coord_sf(datum = NA) +
  theme_void() +
  theme(
    strip.background = element_rect(fill = "white", colour = "black"),
    strip.text       = element_text(face = "bold", size = 12),
    legend.position  = "right",
    plot.title       = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle    = element_text(size = 10, hjust = 0.5),
    plot.caption     = element_text(size = 8, hjust = 1)
  ) +
  labs(
    title    = "Figure 2. Five_33Variable Hardship Index by County (2018–2022)",
    subtitle = "Higher z_scores indicate greater socioeconomic + food‑access stress",
    caption  = "Source: CHR · USDA · FCC"
  )
3
```

# Bivariate Relationships
Figure 2. Time Series of Mortality and Key Indices
• Line plot with year on the x axis, overlaid curves for:
•	Age adjusted mortality
•	Hardship index
•	Broadband coverage (%)
• Facet by county or use multiple y axes; shows parallel trends and inflection post 2020.


```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)    # for rescale()

# 1. Build a trends table from all_years:
df_trends <- all_years %>%
  transmute(
    county      = county,
    year        = year,
    mortality   = mortality_rate,      # raw mortality
    hardship    = hardship_index_5,    # your 5‑var index
    broadband   = pct_broadband        # percent ≥3 providers
  ) %>%
  group_by(county) %>%
  # rescale each series to [0,1] within each county
  mutate(
    mort_std     = rescale(mortality, to = c(0,1)),
    hardship_std = rescale(hardship,  to = c(0,1)),
    bb_std       = rescale(broadband, to = c(0,1))
  ) %>%
  ungroup() %>%
  pivot_longer(
    cols      = ends_with("_std"),
    names_to  = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = recode(metric,
      mort_std     = "Mortality",
      hardship_std = "Hardship",
      bb_std       = "Broadband")
  )

# 2. Plot Figure 2:
ggplot(df_trends, aes(x = year, y = value, color = metric)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ county, ncol = 2) +
  scale_color_manual(
    values = c(
      "Mortality" = "#222222",
      "Hardship"  = "#2ca02c",
      "Broadband" = "#1f77b4"
    )
  ) +
  scale_x_continuous(breaks = 2018:2022) +
  labs(
    title    = "Figure 3###. Time Series of Mortality, Hardship & Broadband (2018–2022)",
    subtitle = "Values rescaled 0–1 within each county for comparability",
    x        = "Year",
    y        = "Rescaled Metric (0 = lowest observed, 1 = highest)",
    color    = NULL,
    caption  = "Source: PA Dept. of Health, CHR, USDA, FCC"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position   = "bottom",
    strip.text        = element_text(face = "bold"),
    plot.title        = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle     = element_text(size = 10, hjust = 0.5)
  )

```

```{r}
#── Libraries ─────────────────────────────────────────────────────────────────────
library(dplyr)
library(ggplot2)
library(ggpubr)    # for stat_cor()
library(patchwork) # to combine plots

#── Prepare data ─────────────────────────────────────────────────────────────────
df_scatter <- all_years %>%
  transmute(
    county    = county,
    year      = year,
    mortality = mortality_rate,
    hardship  = hardship_index_5,
    food_lim  = percent_limited_access,
    broadband = pct_broadband,
    period    = if_else(year < 2020, "Pre‑COVID", "Post‑COVID")
  )

#── Helper to build one scatter panel ──────────────────────────────────────────────
make_plot <- function(x, xlab) {
  ggplot(df_scatter, aes_string(x = x, y = "mortality")) +
    geom_point(alpha = 0.7) +
    geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
    stat_cor(method = "pearson",
             label.x.npc = "left", label.y.npc = 0.95,
             size = 3) +
    labs(x = xlab, y = "Age‑adjusted mortality\n(per 100 000)") +
    theme_minimal(base_size = 12)
}

#── A. Panel per Year ───────────────────────────────────────────────────────────────
p1 <- make_plot("hardship",  "Hardship Index (z)")   + facet_wrap(~year)
p2 <- make_plot("food_lim",  "% Limited Access")     + facet_wrap(~year)
p3 <- make_plot("broadband","% ≥3 Broadband Providers") + facet_wrap(~year)

figure3A <- (p1 | p2 | p3) +
  plot_annotation(
    title = "Figure 3A. Mortality vs. Key Predictors (Faceted by Year)",
    caption = "Points = counties; lines = OLS fits; r = Pearson correlation"
  )

#── B. Facet Pre vs Post COVID ─────────────────────────────────────────────────────
q1 <- make_plot("hardship",  "Hardship Index (z)")   + facet_wrap(~period)
q2 <- make_plot("food_lim",  "% Limited Access")     + facet_wrap(~period)
q3 <- make_plot("broadband","% ≥3 Broadband Providers") + facet_wrap(~period)

figure3B <- (q1 | q2 | q3) +
  plot_annotation(
    title = "Figure 3B. Mortality vs. Predictors (Pre‑ vs Post‑COVID)",
    caption = "Each panel shows OLS trend + Pearson _r_"
  )

#── Print both variants ────────────────────────────────────────────────────────────
figure3A
figure3B

```

# Figure 4. Correlation Heatmap
• Pearson (and/or Spearman) correlations among all continuous variables
• Highlights multicollinearity and guides your index construction story.

```{r}
#── Libraries ─────────────────────────────────────────────────────────────────────
library(dplyr)
library(reshape2)   # for melt()
library(ggplot2)
library(gridExtra)  # to arrange two plots side by side

#── 1. Select & prepare variables ─────────────────────────────────────────────────
# From your stacked all_years data frame:
vars_df <- all_years %>%
  transmute(
    Mortality    = mortality_rate,
    Hardship     = hardship_index_5,
    FoodAccess   = percent_food_insecure,
    LimAccess    = percent_limited_access,
    Broadband    = pct_broadband
  )

#── 2. Compute Correlation Matrices ────────────────────────────────────────────────
pearson_mat <- cor(vars_df, use = "pairwise.complete.obs", method = "pearson")
spearman_mat<- cor(vars_df, use = "pairwise.complete.obs", method = "spearman")

# Melt for ggplot
pearson_melt <- melt(pearson_mat, varnames = c("Var1","Var2"), value.name = "r")
spearman_melt<- melt(spearman_mat, varnames = c("Var1","Var2"), value.name = "r")

#── 3. Plot Helper ─────────────────────────────────────────────────────────────────
plot_corr <- function(melted, title) {
  ggplot(melted, aes(x = Var2, y = Var1, fill = r)) +
    geom_tile(color = "white") +
    geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
    scale_fill_gradient2(
      low    = "steelblue",
      mid    = "white",
      high   = "darkred",
      midpoint = 0,
      limits = c(-1, 1),
      name   = "r"
    ) +
    coord_fixed() +
    theme_minimal(base_size = 12) +
    theme(
      axis.text.x      = element_text(angle = 45, vjust = 1, hjust = 1),
      axis.title       = element_blank(),
      panel.grid       = element_blank(),
      plot.title       = element_text(face = "bold", hjust = 0.5)
    ) +
    labs(title = title)
}

p1 <- plot_corr(pearson_melt,  "Pearson Correlation")
p2 <- plot_corr(spearman_melt, "Spearman Correlation")

#── 4. Display Side by Side ────────────────────────────────────────────────────────
grid.arrange(p1, p2, ncol = 2,
             top = textGrob("Figure 4. Correlation Heatmaps of Key Variables", 
                            gp = gpar(fontsize = 15, fontface = "bold")))

```


```{r}
library(broom)
library(knitr)
#Table 4. Yearly Pearson Correlations

# Compute r and p-value per year for each predictor
table4 <- all_years %>%
  group_by(year) %>%
  summarise(
    r_hardship = cor.test(hardship_index_5, mortality_rate)$estimate %>% unname(),
    p_hardship = cor.test(hardship_index_5, mortality_rate)$p.value,
    r_foodlim  = cor.test(percent_limited_access, mortality_rate)$estimate %>% unname(),
    p_foodlim  = cor.test(percent_limited_access, mortality_rate)$p.value,
    r_bb       = cor.test(pct_broadband, mortality_rate)$estimate %>% unname(),
    p_bb       = cor.test(pct_broadband, mortality_rate)$p.value
  ) %>%
  mutate(across(starts_with("r_"), ~ round(.x, 2)),
         across(starts_with("p_"), ~ format.pval(.x, digits = 2))) %>%
  rename(
    Year                 = year,
    `r (Hardship)`       = r_hardship,
    `p (Hardship)`       = p_hardship,
    `r (Limited Access)` = r_foodlim,
    `p (Limited Access)` = p_foodlim,
    `r (Broadband)`      = r_bb,
    `p (Broadband)`      = p_bb
  )

kable(
  table4,
  caption = "Table 4. Pearson Correlations between Mortality and Key Predictors, by Year"
)

```

```{r}
library(dplyr)
library(broom)
library(purrr)
library(knitr)

#Table 5. Cross-Sectional Regression by Year

# Function to run Model A and Model B for one year
run_year_models <- function(df, yr) {
  d <- filter(df, year == yr)
  mA <- lm(mortality_rate ~ hardship_index_3, data = d)
  mB <- lm(mortality_rate ~ hardship_index_3 + percent_limited_access, data = d)
  
  tibble(
    Year      = yr,
    Model     = c("A", "B"),
    Predictor = c("Hardship", "Limited Access"),
    Estimate  = c(coef(mA)["hardship_index_3"], coef(mB)["percent_limited_access"]),
    conf.low  = c(confint(mA)["hardship_index_3",1], confint(mB)["percent_limited_access",1]),
    conf.high = c(confint(mA)["hardship_index_3",2], confint(mB)["percent_limited_access",2]),
    p.value   = c(summary(mA)$coefficients["hardship_index_3","Pr(>|t|)"],
                  summary(mB)$coefficients["percent_limited_access","Pr(>|t|)"])
  )
}

# Apply across years
table5 <- map_dfr(2019:2022, ~ run_year_models(all_years, .x)) %>%
  filter(!(Year == 2018 & Model == "B")) %>%  # if you only reported 2019 & 2022
  mutate(
    Estimate = round(Estimate, 0),
    conf.low = round(conf.low, 0),
    conf.high= round(conf.high, 0),
    p.value  = format.pval(p.value, digits = 2),
    `95% CI` = paste0("(", conf.low, "–", conf.high, ")")
  ) %>%
  select(Year, Model, Predictor, Estimate, `95% CI`, `p` = p.value)

kable(
  table5,
  caption = "Table 5. Cross-Sectional Regression Coefficients (Deaths/100 000) by Year"
)

```

```{r}
# Inspect your panel data
names(df_panel)
str(df_panel)

# If df_panel doesn’t have mortality, try inspecting all_years
names(all_years)
str(all_years)

```


#Table 6. Fixed-Effects Panel Regression
```{r}
library(plm)
library(broom)
library(dplyr)
library(knitr)

# Prepare the panel data
panel_df <- all_years %>%
  select(
    county,
    year,
    mortality = mortality_rate,
    hardship = hardship_index_5,
    food_access = percent_limited_access,
    broadband = pct_broadband
  )

# Fit the within‐county fixed‐effects model
fe_mod <- plm(
  mortality ~ hardship + food_access + broadband,
  data   = panel_df,
  index  = c("county","year"),
  model  = "within"
)

# Tidy & format for the table
table6 <- tidy(fe_mod, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  transmute(
    Predictor = recode(term,
                       hardship    = "Composite Hardship (SD)",
                       food_access = "Percent Limited Access",
                       broadband   = "Broadband Coverage"),
    Estimate  = round(estimate,    0),
    `95% CI`  = paste0("(", round(conf.low,0), "–", round(conf.high,0), ")"),
    p         = format.pval(p.value, digits = 2)
  )

kable(
  table6,
  caption = "Table 6. Fixed-Effects Panel Regression Results (2018–2022)"
)


```

# Table 7. Interaction Effects (Hardship × Post-COVID, etc.)
```{r}
library(dplyr)
library(broom)
library(knitr)

# Add post_covid indicator
panel_df2 <- panel_df %>%
  mutate(post_covid = if_else(year >= 2020, 1, 0))

# Fit interactions in one linear model
int_mod <- lm(
  mortality ~ hardship * post_covid +
               food_access * post_covid +
               broadband * post_covid,
  data = panel_df2
)

# Extract just the interaction terms
table7 <- tidy(int_mod, conf.int = TRUE) %>%
  filter(grepl(":post_covid", term)) %>%
  transmute(
    Predictor = case_when(
      term == "hardship:post_covid"    ~ "Δ Hardship × Post-COVID",
      term == "food_access:post_covid" ~ "Δ Limited Access × Post-COVID",
      term == "broadband:post_covid"   ~ "Δ Broadband × Post-COVID"
    ),
    Estimate = round(estimate, 1),
    `95% CI` = paste0("(", round(conf.low,1), "–", round(conf.high,1), ")"),
    p        = format.pval(p.value, digits = 2)
  )

kable(
  table7,
  caption = "Table 7. Interaction Effects: Post-COVID Modifications"
)

```

